Temporal stability in human interaction networks\tnoteref{mytitlenote}

### 2017-03-18 physicaa/elsarticle-template.tex
adicionado arquivo final gerado pelo Physica A

\usepackage{xr}
\externaldocument{../supportingInformation}

@@

\author[add3]{Deborah Christina Antunes}
\ead{deborahantunes@gmail.com}
\address[add3]{Psychology Course, Federal University of Ceará (UFC), Sobral Campus}

@@

\begin{abstract}
This paper reports on stable (or invariant) properties of human interaction networks,
with benchmarks derived from public email lists. Activity, recognized through messages sent,
along time and topology were observed in snapshots in a timeline, 
and at different scales. Our analysis shows that activity is practically the same for all networks across timescales ranging from seconds to months.
The principal components of the participants in the topological metrics space remain practically unchanged as different sets of messages are considered.
The activity of participants follows the expected scale-free trace, thus yielding the hub, intermediary and peripheral classes of vertices by comparison against the Erd\"os-R\'enyi model. The relative sizes of these three sectors are essentially the same for all email lists and the same along time.
Typically, $<15\%$ of the vertices are hubs, 15-45\% are intermediary and $>45\%$ are peripheral vertices. Similar results for the distribution of participants in the three sectors and for the relative importance of the topological metrics were obtained for 12 additional networks from Facebook, Twitter and Participabr. These properties are consistent with the literature and may be general for human interaction networks, which has important implications for establishing a typology of participants based on quantitative criteria.
\end{abstract}

@@

In this paper we analyze the evolution of human interaction networks.
Directed and weighted representations were built through the observation of replies as links.
Interaction networks from email lists were the most convenient for deriving results and for benchmarking while networks from Facebook, Twitter and Participabr were used for the sake of generalization.
Using a timeline of activity snapshots with a constant number of contiguous messages, we found remarkable stability (or invariance) for important network properties. For instance, activity along different timescales follows specific patterns; the most basic topological metrics can always be combined into characteristic principal components; and the fractions of participants in different sectors do not vary with time. This is not an intuitive result, given that participants constantly transition in network structure. Because these properties were shared by networks from various sources, and are consistent with the literature in complex networks~\cite{newmanBook}, we advocate that the conclusions might be valid for general classes of interaction networks. In particular, this allows us to 
bridge the gap between data analysis and social sciences in the discussion of types of networks and of participants.
It is worth noting that typologies are the canon of scientific literature for the classification of human agents, with pragmatic standards~\cite{myers} and critical paradigms~\cite{adorno,typCanon}.

This paper is organized as follows. Section~\ref{sec:related} describes related work, while data, scripts and methods of analysis are given in Section~\ref{sec:data} and Section~\ref{sec:carac}.
Section~\ref{sec:results} reports results and discussion, leading to Section~\ref{sec:conc} for conclusions.
Supplementary data analysis, including directions for video and sound mappings of network structures, and numeric detailed results for networks from Twitter, Facebook and Participabr, are provided in the Supporting Information document.

@@

We observed various email lists and selected four of them together with data from Twitter, Facebook and Participabr for a thorough analysis,
from which general properties can be inferred. These lists are as follows:

@@

The first 20,000 messages of each list were considered, with basic attributes of total timespan, authors, threads and missing messages indicated in Table~\ref{tab:genLists}. We considered 140 additional email lists to report on the interdependence between the number of participants and the number of discussion threads. Furthermore, 12 networks from Facebook (8), Twitter (2) and Participabr (2) were scrutinized, and their analysis is given in the Supporting Information document for the purpose of testing the generality of the results.

\begin{table}
\centering
	\caption{{\bf Overview of the email lists analyzed.} Columns $date_1$ and $date_M$ have dates of first and last messages from the 20,000 messages considered in each email list.
$N$ is the number of participants (number of different email addresses),
$\Gamma$ is the number of discussion threads (count of messages without antecedent),
$\overline{M}$ is the number of messages missing in the 20,000 collection
($100\frac{23}{20000}=0.115$ percent in the worst case).
}
\label{tab:genLists}
\begin{tabular}{l|c|c|c|c|c}\hline
list & $date_1$ & $date_{M}$    & $N$  & $\Gamma$ & $\overline{M}$ \\\hline
\input{tables/tab1Overview}
\end{tabular}
\end{table}

The data and scripts used to derive the results, figures and tables, and this article itself are publicly available. Email messages are downloadable from the Gmane public database~\cite{Gmanewikipedia}.
Data annotated from Facebook and Twitter are in a public repository~\cite{fbtwData}.
Data from Participabr were used from the linked data/semantic web RDF triples~\cite{opa}, available in~\cite{datahub}.
Computer scripts are delivered through a public domain Python PyPI package and an open Git repository~\cite{gmanePack}.
This open approach to both data and scripts reinforces the scientific aspect of the contribution~\cite{openSci} and mitigates ethical and moral issues involved in researching systems constituted of human individuals~\cite{anPhy,ccs15}.

@@

\begin{table}
	\caption{{\bf Time-related circular statistics.} The rescaled circular mean $\theta_\mu'$ and the circular dispersion $\delta(z)$, described in Section~\ref{sec:mtime}, for different timescales. This example table was constructed using all LAD messages, and the results are the same for other lists, as shown in Section~\ref*{si:circ} of the Supporting Information document. The most uniform distribution of activity was found in seconds and minutes. 	Hours of the day exhibited the most concentrated activity (lowest $\delta(z)$), with mean between 2 p.m. and 3 p.m. ($\theta'=-9.61$). Weekdays, days of the month and months have mean near zero (i.e. near the beginning of the week, month and year) and high dispersion. Note that $\theta_u'$ has the dimensional unit of the corresponding time period while $\delta(z)$ is dimensionless.}
\begin{center}
\begin{tabular}{ l|| c|c }
\hline
%scale & $\theta_\mu'$ & $S(z)$ & $Var(z)$ & $\delta(z)$ & $\frac{max(incidence)}{min(incidence)}$ & $ \mu_{\frac{max(incidence')}{min(incidence')}} $ & $ \sigma_{\frac{max(incidence')}{min(incidence')} } $ \\ \hline\hline
%& $\theta_\mu'$ & $S(z)$ & $Var(z)$ & $\delta(z)$  \\ \hline\hline
scale & mean $\theta_\mu'$ & dispersion $\delta(z)$  \\ \hline
\input{tables/tab2TimeLAD___}
\end{tabular}
\end{center}
\label{tab:circ}
\end{table}

\begin{table}
	\caption{{\bf Activity percentages along the hours of the day.} Nearly identical distributions were observed on other social systems as shown in Section~\ref*{si:hours} of the Supporting Information document.
Highest activity was observed between noon and 6pm (with 1/3 of total day activity), followed by the time period between 6pm and midnight.
Around 2/3 of the activity takes place from noon to midnight
but the activity peak occurs between 11 a.m. and 12 p.m.
This table shows results for the activity in CPP.}
\footnotesize
\input{tables/tabHoursCPP_}
\label{tab:hin}
\end{table}

\begin{table}
	\caption{{\bf Activity percentages along weekdays.}
Higher activity was observed during workweek days, with a decrease of activity on weekend days of at least one third and at most two thirds.}
\begin{center}
\begin{tabular}{ l ||  c | c | c | c | c |   c | c}
\hline
& Mon & Tue & Wed & Thu & Fri & Sat & Sun  \\ \hline
\input{tables/tabWeekdays}
\end{tabular}
\end{center}
\label{tab:win}
\end{table}

@@

\begin{table}
	\caption{{\bf Activity along the days of the month.}
Nearly identical distributions are found in all systems
as indicated in Section~\ref*{si:monthdays} of the Supporting Information. Although slightly higher activity rates are found in the beginning of the month, the most important feature seems to be the homogeneity made explicit by the high circular dispersion in Table~\ref{tab:circ}.
This specific example and empirical table correspond to the activity of the MET email list.}
\footnotesize
\input{tables/tabMonthdaysMET}
\label{tab:min}
\end{table}

\begin{table}
	\caption{{\bf Activity percentages on months along the year.} 	Activity is usually concentrated in Jun-Aug and/or in Dec-Mar, potentially due to academic calendars, vacations and end-of-year holidays. This table corresponds to activity in LAU. Similar results are shown for other lists in Section~\ref*{si:months} of the Supporting Information document.}
\footnotesize
\input{tables/tabMonthsLAU}
\label{tab:min2}
\end{table}

@@

In order to support the generality of these findings,
we list the Erd\"os sector sizes of 12 networks from Facebook, Twitter and Participabr in Table~\ref*{tab:secE} of the Supporting Information document. The fractions of hubs, intermediary and periphery nodes are
essentially the same as for the email list networks but with exceptions and a greater variability.

\begin{figure*} 
\centering
\includegraphics[width=\textwidth]{figs/InText-WLAU-S1000__}
	\caption{{\bf Stability of Erd\"os sector sizes.}
Fractions of participants derived from degree and strength criteria, $E_1$ and $E_4$ described in Section~\ref{sectioning}, are both on the left.
Fractions derived from the exclusivist $C_1$ and the inclusivist $C_2$ compound criteria are shown in the plots to the right.
The ordinates $\overline{e_{\gamma,\phi}}=\frac{|e_{\gamma,\phi}|}{N}$ denote the fraction of participants in sector $\phi$ through criterion $E_\gamma$
and, similarly, $\overline{c_{\delta,\phi}}=\frac{|c_{\delta,\phi}|}{N}$ denotes the fraction of participants in sector $\phi$ through criterion $C_\delta$.
Sections~\ref*{si:frac} and~\ref*{si:ext} of the Supporting Information bring a systematic collection of such timeline figures with all simple and compound criteria specified in Section~\ref{sectioning}, with results for networks from Facebook, Twitter and Participabr.}
\label{fig:sectIL}
\end{figure*}

\begin{table}[h]
	\caption{{\bf Distribution of activity among participants.}
The first column shows the percentage of messages sent by the most active participant. The column for the first quartile ($Q_1$) gives the minimum percentage of participants responsible for at least 25\% of total messages with the actual percentage in parentheses. Similarly, the column for the first three quartiles $Q_3$ gives the minimum percentage of participants responsible for 75\% of total messages.
The last decile $D_{-1}$ column shows the maximum percentage of participants responsible for 10\% of messages.}
\begin{center}
\begin{tabular}{  l ||  c | c | c | c  }
\hline
list & hub & $ Q_1 $ & $ Q_3 $ & $D_{-1}$ \\ \hline
\input{tables/userTab}
\end{tabular}
\end{center}
\label{autores}
\end{table}

@@

\begin{table}[!h]
	\caption{{\bf Invariance of principal components.} Loadings for the 14 metrics into the principal components for the MET list, $1000$ messages in 20 disjoint positions. The clustering coefficient (cc) appears as the first metric in the table, followed by 7 centrality metrics and 6 symmetry-related metrics. Note that the centrality measurements, including degrees, strength and betweenness centrality, are the most important contributors for the first principal component, while the second component is dominated by symmetry metrics. The clustering coefficient is only relevant for the third principal component. The three components have in average more than 85\% of the variance.
The low standard deviation $\sigma$ implies that the principal components are considerably stable.}
\footnotesize
\input{tables/tabPCA3CPP}
\label{tab:pcain}
\end{table}

@@

\section{Conclusions}\label{sec:conc}
The very small standard deviations of principal components formation
(see Sections~\ref{sec:pca} and~\ref{prevalence}),
the presence of the Erd\"os sectors even in networks with
few participants (see Sections~\ref{sectioning} and~\ref{subsec:pih}),
and the recurrent activity patterns along different timescales (see Sections~\ref{sec:mtime} and~\ref{constDisc}),
go a step further in characterizing scale-free networks in the context
of the interaction of human individuals.
Furthermore, the importance of symmetry-related metrics,
which surpassed that of clustering coefficient,
with respect to dispersion of the system in the topological measures space,
might add to the current understanding of key-differences between digraphs and
undirected graphs in complex networks.
Noteworthy is also the very stable fraction participants in each Erd\"os sector when the network reaches more than 200 participants.
Benchmarks were derived from email list networks
and the supplied analysis of
networks from Facebook,
Twitter and Participabr in the Supporting Information might ease hypothesizing
about the generality of these characteristics.

@@

\subsection{Acknowledgments}
Financial support was obtained from CNPq (140860/2013-4,
project 870336/1997-5), United Nations Development Program (contract: 2013/000566; project BRA/12/018) and FAPESP. 
The authors are grateful to the American Jewish Committee for maintaining an online copy of the Adorno book used on the epigraph~\cite{adorno}, to Gmane creators and maintainers for the public email list data, to the communities of the email lists and other groups used in the analysis, and to the Presidency of the Brazilian Republic for keeping Participabr code and data open.
We are also grateful to developers and users of Python scientific tools, to Leonardo Paulo Maia (IFSC/USP) and to Francisco J. P. Lopes (UFRJ) for valuable insights.

@@

\bibliography{../paper}

>>>>

\usepackage{xr}
\externaldocument{supportingInformation}

@@

\author[add3]{Deborah Christina Antunes}
\ead{deborahantunes@gmail.com}
\address[add3]{Psychology Course, Federal University of Cear\'a (UFC), Sobral Campus}

@@

\begin{abstract}
This paper reports on stable (or invariant) properties of human interaction networks,
with benchmarks derived from public email lists. Activity, recognized through messages sent,
along time and topology were observed in snapshots in a timeline, 
and at different scales. Our analysis shows that activity is practically the same for all networks across timescales ranging from seconds to months.
The principal components of the participants in the topological metrics space remain practically unchanged as different sets of messages are considered.
The activity of participants follows the expected scale-free trace, thus yielding the hub, intermediary and peripheral classes of vertices by comparison against the Erd\"os-R\'enyi model. The relative sizes of these three sectors are essentially the same for all email lists and the same along time.
Typically, $<15\%$ of the vertices are hubs, 15-45\% are intermediary and $>45\%$ are peripheral vertices. Similar results for the distribution of participants in the three sectors and for the relative importance of the topological metrics were obtained for 12 additional networks from Facebook, Twitter and ParticipaBR. These properties are consistent with the literature and may be general for human interaction networks, which has important implications for establishing a typology of participants based on quantitative criteria.
\end{abstract}

@@

In this paper we analyze the evolution of human interaction networks.
Directed and weighted representations were built through the observation of replies as links.
Interaction networks from email lists were the most convenient for deriving results and for benchmarking while networks from Facebook, Twitter and ParticipaBR were used for the sake of generalization.
Using a timeline of activity snapshots with a constant number of contiguous messages, we found remarkable stability (or invariance) for important network properties. For instance, activity along different timescales follows specific patterns; the most basic topological metrics can always be combined into characteristic principal components; and the fractions of participants in different sectors do not vary with time. This is not an intuitive result, given that participants constantly transition in network structure. Because these properties were shared by networks from various sources, and are consistent with the literature in complex networks~\cite{newmanBook}, we advocate that the conclusions might be valid for general classes of interaction networks. In particular, this allows us to 
bridge the gap between data analysis and social sciences in the discussion of types of networks and of participants.
It is worth noting that typologies are the canon of scientific literature for the classification of human agents, with pragmatic standards~\cite{myers} and critical paradigms~\cite{adorno,typCanon}.

This paper is organized as follows. Section~\ref{sec:related} describes related work, while data, scripts and methods of analysis are given in Section~\ref{sec:data} and Section~\ref{sec:carac}.
Section~\ref{sec:results} reports results and discussion, leading to Section~\ref{sec:conc} for conclusions.
Supplementary data analysis, including directions for video and sound mappings of network structures, and numeric detailed results for networks from Twitter, Facebook and ParticipaBR, are provided in the Supporting Information document.

@@

We observed various email lists and selected four of them together with data from Twitter, Facebook and ParticipaBR for a thorough analysis,
from which general properties can be inferred. These lists are as follows:

@@

The first 20,000 messages of each list were considered, with basic attributes of total timespan, authors, threads and missing messages indicated in Table~\ref{tab:genLists}. We considered 140 additional email lists to report on the interdependence between the number of participants and the number of discussion threads. Furthermore, 12 networks from Facebook (8), Twitter (2) and ParticipaBR (2) were scrutinized, and their analysis is given in the Supporting Information document for the purpose of testing the generality of the results.

\begin{table}
\centering
\label{tab:genLists}
\begin{tabular}{l|c|c|c|c|c}\hline
list & $date_1$ & $date_{M}$    & $N$  & $\Gamma$ & $\overline{M}$ \\\hline
\input{tables/tab1Overview}
\end{tabular}
	\caption{{\bf Overview of the email lists analyzed.} Columns $date_1$ and $date_M$ have dates of first and last messages from the 20,000 messages considered in each email list.
$N$ is the number of participants (number of different email addresses),
$\Gamma$ is the number of discussion threads (count of messages without antecedent),
$\overline{M}$ is the number of messages missing in the 20,000 collection
($100\frac{23}{20000}=0.115$ percent in the worst case).
}
\end{table}

The data and scripts used to derive the results, figures and tables, and this article itself are publicly available. Email messages are downloadable from the Gmane public database~\cite{Gmanewikipedia}.
Data annotated from Facebook and Twitter are in a public repository~\cite{fbtwData}.
Data from ParticipaBR were used from the linked data/semantic web RDF triples~\cite{opa}, available in~\cite{datahub}.
Computer scripts are delivered through a public domain Python PyPI package and an open Git repository~\cite{gmanePack}.
This open approach to both data and scripts reinforces the scientific aspect of the contribution~\cite{openSci} and mitigates ethical and moral issues involved in researching systems constituted of human individuals~\cite{anPhy,ccs15}.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l|| c|c }
\hline
%scale & $\theta_\mu'$ & $S(z)$ & $Var(z)$ & $\delta(z)$ & $\frac{max(incidence)}{min(incidence)}$ & $ \mu_{\frac{max(incidence')}{min(incidence')}} $ & $ \sigma_{\frac{max(incidence')}{min(incidence')} } $ \\ \hline\hline
%& $\theta_\mu'$ & $S(z)$ & $Var(z)$ & $\delta(z)$  \\ \hline\hline
scale & mean $\theta_\mu'$ & dispersion $\delta(z)$  \\ \hline
\input{tables/tab2TimeLAD___}
\end{tabular}
\end{center}
	\caption{{\bf Time-related circular statistics.} The rescaled circular mean $\theta_\mu'$ and the circular dispersion $\delta(z)$, described in Section~\ref{sec:mtime}, for different timescales. This example table was constructed using all LAD messages, and the results are the same for other lists, as shown in Section~\ref*{si:circ} of the Supporting Information document. The most uniform distribution of activity was found in seconds and minutes. 	Hours of the day exhibited the most concentrated activity (lowest $\delta(z)$), with mean between 2 p.m. and 3 p.m. ($\theta'=-9.61$). Weekdays, days of the month and months have mean near zero (i.e. near the beginning of the week, month and year) and high dispersion. Note that $\theta_u'$ has the dimensional unit of the corresponding time period while $\delta(z)$ is dimensionless.}
\label{tab:circ}
\end{table}

\begin{table}
\footnotesize
\input{tables/tabHoursCPP_}
\label{tab:hin}
	\caption{{\bf Activity percentages along the hours of the day.} Nearly identical distributions were observed on other social systems as shown in Section~\ref*{si:hours} of the Supporting Information document.
Highest activity was observed between noon and 6pm (with 1/3 of total day activity), followed by the time period between 6pm and midnight.
Around 2/3 of the activity takes place from noon to midnight
but the activity peak occurs between 11 a.m. and 12 p.m.
This table shows results for the activity in CPP.}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{ l ||  c | c | c | c | c |   c | c}
\hline
& Mon & Tue & Wed & Thu & Fri & Sat & Sun  \\ \hline
\input{tables/tabWeekdays}
\end{tabular}
\end{center}
	\caption{{\bf Activity percentages along weekdays.}
Higher activity was observed during workweek days, with a decrease of activity on weekend days of at least one third and at most two thirds.}
\label{tab:win}
\end{table}

@@

\begin{table}
\footnotesize
\input{tables/tabMonthdaysMET}
\label{tab:min}
	\caption{{\bf Activity along the days of the month.}
Nearly identical distributions are found in all systems
as indicated in Section~\ref*{si:monthdays} of the Supporting Information. Although slightly higher activity rates are found in the beginning of the month, the most important feature seems to be the homogeneity made explicit by the high circular dispersion in Table~\ref{tab:circ}.
This specific example and empirical table correspond to the activity of the MET email list.}
\end{table}

\begin{table}
\footnotesize
\input{tables/tabMonthsLAU}
\label{tab:min2}
	\caption{{\bf Activity percentages on months along the year.} 	Activity is usually concentrated in Jun-Aug and/or in Dec-Mar, potentially due to academic calendars, vacations and end-of-year holidays. This table corresponds to activity in LAU. Similar results are shown for other lists in Section~\ref*{si:months} of the Supporting Information document.}
\end{table}

@@

In order to support the generality of these findings,
we list the Erd\"os sector sizes of 12 networks from Facebook, Twitter and ParticipaBR in Table~\ref*{tab:secE} of the Supporting Information document. The fractions of hubs, intermediary and periphery nodes are
essentially the same as for the email list networks but with exceptions and a greater variability.

\begin{figure*} 
\centering
\includegraphics[width=\textwidth]{figs/InText-WLAU-S1000__}
	\caption{{\bf Stability of Erd\"os sector sizes.}
Fractions of participants derived from degree and strength criteria, $E_1$ and $E_4$ described in Section~\ref{sectioning}, are both on the left.
Fractions derived from the exclusivist $C_1$ and the inclusivist $C_2$ compound criteria are shown in the plots to the right.
The ordinates $\overline{e_{\gamma,\phi}}=\frac{|e_{\gamma,\phi}|}{N}$ denote the fraction of participants in sector $\phi$ through criterion $E_\gamma$
and, similarly, $\overline{c_{\delta,\phi}}=\frac{|c_{\delta,\phi}|}{N}$ denotes the fraction of participants in sector $\phi$ through criterion $C_\delta$.
Sections~\ref*{si:frac} and~\ref*{si:ext} of the Supporting Information bring a systematic collection of such timeline figures with all simple and compound criteria specified in Section~\ref{sectioning}, with results for networks from Facebook, Twitter and ParticipaBR.}
\label{fig:sectIL}
\end{figure*}

\begin{table}[h]
\begin{center}
\begin{tabular}{  l ||  c | c | c | c  }
\hline
list & hub & $ Q_1 $ & $ Q_3 $ & $D_{-1}$ \\ \hline
\input{tables/userTab}
\end{tabular}
\end{center}
	\caption{{\bf Distribution of activity among participants.}
The first column shows the percentage of messages sent by the most active participant. The column for the first quartile ($Q_1$) gives the minimum percentage of participants responsible for at least 25\% of total messages with the actual percentage in parentheses. Similarly, the column for the first three quartiles $Q_3$ gives the minimum percentage of participants responsible for 75\% of total messages.
The last decile $D_{-1}$ column shows the maximum percentage of participants responsible for 10\% of messages.}
\label{autores}
\end{table}

@@

\begin{table}[!h]
\footnotesize
\input{tables/tabPCA3CPP}
\label{tab:pcain}
	\caption{{\bf Invariance of principal components.} Loadings for the 14 metrics into the principal components for the MET list, $1000$ messages in 20 disjoint positions. The clustering coefficient (cc) appears as the first metric in the table, followed by 7 centrality metrics and 6 symmetry-related metrics. Note that the centrality measurements, including degrees, strength and betweenness centrality, are the most important contributors for the first principal component, while the second component is dominated by symmetry metrics. The clustering coefficient is only relevant for the third principal component. The three components have in average more than 85\% of the variance.
The low standard deviation $\sigma$ implies that the principal components are considerably stable.}
\end{table}

@@

\section{Conclusions}\label{sec:conc}
The very small standard deviations of principal components formation
(see Sections~\ref{sec:pca} and~\ref{prevalence}),
the presence of the Erd\"os sectors even in networks with
few participants (see Sections~\ref{sectioning} and~\ref{subsec:pih}),
and the recurrent activity patterns along different timescales (see Sections~\ref{sec:mtime} and~\ref{constDisc}),
go a step further in characterizing scale-free networks in the context
of the interaction of human individuals.
Furthermore, the importance of symmetry-related metrics,
which surpassed that of clustering coefficient,
with respect to dispersion of the system in the topological measures space,
might add to the current understanding of key-differences between digraphs and
undirected graphs in complex networks.
Noteworthy is also the very stable fraction participants in each Erd\"os sector when the network reaches more than 200 participants.
Benchmarks were derived from email list networks
and the supplied analysis of
networks from Facebook,
Twitter and ParticipaBR in the Supporting Information might ease hypothesizing
about the generality of these characteristics.

@@

\subsection{Acknowledgments}
Financial support was obtained from CNPq (140860/2013-4,
project 870336/1997-5), United Nations Development Program (contract: 2013/000566; project BRA/12/018) and FAPESP. 
The authors are grateful to the American Jewish Committee for maintaining an online copy of the Adorno book used on the epigraph~\cite{adorno}, to Gmane creators and maintainers for the public email list data, to the communities of the email lists and other groups used in the analysis, and to the Presidency of the Brazilian Republic for keeping ParticipaBR code and data open.
We are also grateful to developers and users of Python scientific tools, to Leonardo Paulo Maia (IFSC/USP) and to Francisco J. P. Lopes (UFRJ) for valuable insights.

@@

\bibliography{paper}
